An Introduction to Rocker: Docker Containers for R

### 2017-07-10 manuscript.Rmd
second pass with typo edits
minimal style / wording / content changes

**Build schedule**: The official `r-base` image is rebuilt by Docker
following any updates to the official `debian` images (roughly every
few weeks).  The rest of the stack uses build triggers that rebuild the
images either whenever `r-base` is updated or the Dockerfile sources are
updated on the corresponding GitHub repository.  The only exception in
this stack is the `drd` image, which is rebuilt each week by cron trigger.

@@

<!-- Should we mention r-apt separately? -->

<!-- Do we describe version tags of r-base, which do provide older versions of R but are based on Dockerfiles which will no longer build since the requested R version is no longer in `unstable`?  -->

@@

This stack emphasizes stability and reproducibility of the Docker build.
This stack was introduced much more recently, starting in November
of 2016, in response to considerable user input and requests. The key
feature of this stack is the ability to run older versions of R, along
with the contemporaneous versions of R packages.  A user specifies the
version desired using an image tag, e.g.  `rocker/r-ver:3.3.1` will
refer to an image with R version 3.3.1 installed.  Omitting the tag
is equivalent to using the tag `latest`, which, as the name implies,
will always point to an image using the current R release.  Thus, users
wanting to create downstream Dockerfiles which are based on the current
release at time but will continue to reconstruct the same environment in
the future after newer R versions are released should explicilty include
the corresponding version tag, e.g. `rocker/r-ver:3.4.0` at the time
of writing, and not the `latest` tag.  Users can also run the current
development version of R using the tag `devel`, which is built nightly
from R-devel sources from `subversion`.

**MRAN archives**: To facilitate installation of only contemporaneous
versions of R packages on these images, the default CRAN mirror from which
to install R packages is fixed to a snapshot of CRAN corresponding to
the last date for which that version of R was the latest release. These
snapshots are provided by the MRAN archive created by Revolution
Analytics (now aquired by Microsoft), which archives daily snapshots
of all of CRAN, from which a user can install packages with the usual
`install.packages()` function  [@MRAN]. Users can always override this
default by passing any current CRAN repository explicitly.  Unlike CRAN,
Bioconductor only updates its repositories through bi-annual releases
aligned to R's Spring release schedule. Thus, Bioconductor packages
can be installed in the usual way using `bioclite`, which automatically
selects the Bioconductor release corresponding to the version of R in use.

**Version tags**:  These version tags are propegated throughout this
stack: for instance, `rocker/tidyverse:devel` will provide the currently
released versions of the R packages in the `tidyverse` [@tidyverse]
installed on the nightly build of R-devel.  Developers building packages
on this stack are encouraged to tag their images accordingly as well.
Table 3 indicates which versions of R are currently available in the
stack, going back to `3.1.0`.  While older versions may be added to
the stack at a later date, we note that the MRAN snapshots began in
starting 2014-09-17 and thus go back only to R `3.1` era.  Each tag must
be built from a separate Dockerfile, enabling minor differences in the
build instructions to accomidate changing dependencies.  Dockerfiles for
past versions (e.g. < `3.4.0` currently) are intended to remain static
over the long term, while the tag for the current version, `latest`,
and `devel` may be tweaked to accomidate new features or dependencies.

**Installation**: In this stack, the desired version of R is always built
directly from source rather than the `apt` repositories.  Compilers and
dependencies are still installed from the stable `apt` repositories,
and thus lag behind the more recent versions found in the `testing`
stack. Version tags `3.3.3` and older are based on the Debian 8.0 release,
codenamed `jessie`, while `3.4.0`, `devel`, and `latest` are based on
Debian 9.0, `stretch`, released 2017-06-17, and thus have access to much
newer versions of common system dependencies and compilers.  Dependencies
needed to compile R that are not required at runtime are removed once R is
installed, keeping the base images light-weight for faster download times.
While most system dependencies required by common R packages can still
be installed from the `apt` repositories, occassionally a more recent
version must be compiled from source (e.g. the Gibbs Sampling library,
JAGS, and the geospatial toolkit GDAL, must both be compiled from source
on `debian:jessie` images).  In this stack, users should never install R
packages, using `apt`, since this will install a second, older version
of R from the Debian repositories and a dated version of the R package
(since any `r-cran-pkgname` package in the Debian repositories will
depend on `r-base` in `apt` as well).

@@

The `rstudio` image includes a lightweight, easy-to-use and
docker-friendly `init` system, s6 [@s6] for running persistant services,
including the RStudio server.  This system provides a convenient way for
downstream Dockerfile developers to add additional persistent services
(such as an ssh server) to a single container, or additional start-up
or shutdown scripts that should be run when an instance starts up or
shuts down.  The `rstudio` image uses such a start-up script to configure
user settings such as login password and permissions through environmental
variables at run time, see *Using Rocker Images* below.

The `tidyverse` image contains all required and suggested dependencies
of the commonly used `tidyverse` and `devtools` R packages, including
external database libraries (e.g. Maria-DB and Postgresql).  Users should
consult the package Dockerfiles or `installed.packages()` list directly
for a complete list of installed packages.  The `verse` library adds
commonly used dependencies, notably a large but not comprehensive LaTeX
environment and java development libraries.  Previously the Rocker
project provided the image `hadleyverse` which was since divided into
`tidyverse` and `verse` through community input.

@@

Installation of certain R packages can be challenging and non-intuitive,
particularly when those packages require specific external
dependencies. By running R through a Rocker image, users can deploy an
R environment where their desired packages and libraries have already
been installed and configured. As we discuss below, this configuration
benefits from the input of a large community of users, helping to
ensure better and more reliable performance. For instance, one example
of such a configuration is in building and testing R packages using
compiler sanatizers \citep{edd_sanitizers}, which requires R itself
be re-compiled with specialized settings not avaialable or familiar
to most R users.  Using the Rocker images `rocker/r-devel-san` and
`rocker/r-devel-ubsan-clang` provides a fast and simple way for users
to test

>>>>

**Build schedule**: The official `r-base` image is rebuilt by Docker
following any updates to the official `debian` images (roughly every
few weeks).  The rest of the stack uses build triggers that rebuild the
images either whenever `r-base` is updated, or the Dockerfile sources are
updated on the corresponding GitHub repository.  The only exception in
this stack is the `drd` image, which is rebuilt each week by cron trigger.

@@

<!-- CB: Should we mention r-apt separately? -->
<!-- DE: Or maybe not at all. It is somewhat specialised. -->

<!-- CB: Do we describe version tags of r-base, which do provide older versions of R but are based on Dockerfiles which will no longer build since the requested R version is no longer in `unstable`?  -->
<!-- DE: Not sure. -->

@@

This stack emphasizes stability and reproducibility of the Docker build.
This stack was introduced much more recently (November 2016) 
in response to considerable user input and requests. The key
feature of this stack is the ability to run older versions of R along
with the contemporaneous versions of R packages.  A user specifies the
version desired using an image tag, e.g.  `rocker/r-ver:3.3.1` will
refer to an image with R version 3.3.1 installed.  Omitting the tag
is equivalent to using the tag `latest`, which, as the name implies,
will always point to an image using the current R release.  Thus, users
wanting to create downstream Dockerfiles which are based on the current
release at time but will continue to reconstruct the same environment in
the future after newer R versions are released should explicilty include
the corresponding version tag, e.g. `rocker/r-ver:3.4.0` at the time
of writing, and not the `latest` tag.  Users can also run the current
development version of R using the tag `devel`, which is built nightly
from R-devel sources from `subversion`.

**MRAN archives**: To facilitate installation of only contemporaneous
versions of R packages on these images, the default CRAN mirror from which
to install R packages is fixed to a snapshot of CRAN corresponding to
the last date for which that version of R was the latest release. These
snapshots are provided by the MRAN archive created by Revolution
Analytics (now part of Microsoft). It archives daily snapshots
of all of CRAN from which a user can install packages with the usual
`install.packages()` function  [@MRAN]. Users can always override this
default by passing any current CRAN repository explicitly.  Unlike CRAN,
Bioconductor only updates its repositories through bi-annual releases
aligned to R's Spring release schedule. Thus, Bioconductor packages
can be installed in the usual way using `bioclite`, which automatically
selects the Bioconductor release corresponding to the version of R in use.

**Version tags**:  These version tags are propegated throughout this
stack: for instance, `rocker/tidyverse:devel` will provide the currently
released versions of the R packages in the `tidyverse` [@tidyverse]
installed on the nightly build of R-devel.  Developers building packages
on this stack are encouraged to tag their images accordingly as well.
Table 3 indicates which versions of R are currently available in the
stack, going back to `3.1.0`.  While older versions may be added to
the stack at a later date, we note that the MRAN snapshots began in
starting 2014-09-17 and thus go back only to R `3.1` era.  Each tag must
be built from a separate Dockerfile, enabling minor differences in the
build instructions to accomidate changing dependencies.  Dockerfiles for
past versions (e.g. < `3.4.0` currently) are intended to remain static
over the long term, while the tag for the current version, `latest`,
and `devel` may be tweaked to accomodate new features or dependencies.

**Installation**: In this stack, the desired version of R is always built
directly from source rather than the `apt` repositories.  Compilers and
dependencies are still installed from the stable `apt` repositories,
and thus lag behind the more recent versions found in the `testing`
stack. Version tags `3.3.3` and older are based on the Debian 8.0 release,
codenamed `jessie`, while `3.4.0`, `devel`, and `latest` are based on
Debian 9.0, `stretch`, released 2017-06-17, and thus have access to much
newer versions of common system dependencies and compilers.  Dependencies
needed to compile R that are not required at runtime are removed once R is
installed, keeping the base images light-weight for faster download times.
While most system dependencies required by common R packages can still
be installed from the `apt` repositories, occassionally a more recent
version must be compiled from source (e.g. the Gibbs Sampling library,
JAGS, and the geospatial toolkit GDAL, must both be compiled from source
on `debian:jessie` images).  In this stack, users should never install R
packages, using `apt`, since this will install a second, older version
of R from the Debian repositories and a dated version of the R package
(since any `r-cran-pkgname` package in the Debian repositories will
depend on `r-base` in `apt` as well).
<!-- DE: Likely a non-issue as /usr/local is always search before /usr. -->

@@

The `rstudio` image includes a lightweight, easy-to-use and
docker-friendly `init` system, s6 [@s6] for running persistant services,
including the RStudio server.  This system provides a convenient way for
downstream Dockerfile developers to add additional persistent services
(such as an `ssh` server) to a single container, or additional start-up
or shutdown scripts that should be run when an instance starts up or
shuts down.  The `rstudio` image uses such a start-up script to configure
user settings such as login password and permissions through environmental
variables at run time, see *Using Rocker Images* below.

The `tidyverse` image contains all required and suggested dependencies
of the commonly used `tidyverse` and `devtools` R packages, including
external database libraries (e.g. MariaDB and PostgreSQL).  Users should
consult the package Dockerfiles or `installed.packages()` list directly
for a complete list of installed packages.  The `verse` library adds
commonly used dependencies, notably a large but not comprehensive LaTeX
environment and Java development libraries.  Previously the Rocker
project provided the image `hadleyverse` which was since divided into
`tidyverse` and `verse` through community input.

@@

Installation of certain R packages can be challenging,
particularly when those packages require specific external
dependencies. By running R through a Rocker image, users can deploy an
R environment where their desired packages and libraries have already
been installed and configured. As we discuss below, this configuration
benefits from the input of a large community of users, helping to
ensure better and more reliable performance. For instance, one example
of such a configuration is in building and testing R packages using
compiler sanatizers \citep{edd_sanitizers}, which requires R itself
be re-compiled with specialized settings not avaialable or familiar
to most R users.  Using the Rocker images `rocker/r-devel-san` and
`rocker/r-devel-ubsan-clang` provides a fast and simple way for users
to test.
